JD notes on effects

### 2022-03-23 manuscript/vareffects_plosone.tex
Autocommit
population-based > sample-based

In many applications which use generalized linear (mixed) models, outcome predictions or predicted probabilities are often of interest. For models that involve complex multiplicative interactions, additional non-focal predictors or nonlinear link functions, the estimated coefficient are not readily interpretable. A general way to summarize these kind of models is through predictor effects, which are not only dependent on choice of representative values of focal predictor but also sensitive to which values of non-focal predictors are chosen. The most common approach is generating predictor effects at a ``reference point'', usually the mean, of non-focal predictor, i.e., mean-based which could be considered as the effect of an ``average case'' in the population. In the presence of sources of bias such as additional non-focal predictors, nonlinear link functions, random effect terms, etc., mean-based approach generate predictor effects that are biased and may not be consistent with the observed quantities. An alternative is the population-based approach which estimates the average effect in the population. Moreover, isolated confidence intervals provide an alternative and a more clear way to describe uncertainty associated with the focal predictor of interest. In addition to theoretical and methodical comparison, using simulation, we illustrate the two approaches and show that they can produce substantially different results and that population-based approach can not only produce estimates consistent with the observed values, but also appropriate for bias correction. We also present an alternative way, isolated confidence intervals, to describe uncertainties associated with these predictions.

@@

When dealing with nonlinear link functions or models with multiple non-focal predictors, the correct predictions, for example, are even much harder to estimate. One approach is to make predictions on the transformed scale (linear predictor scale), and then back-transform to the original scale. However, the back-transformation may either result in biased predictions or requires some approximation. In particular, bias in expected mean prediction induced by nonlinear transformation of the response variable can lead inaccurate predictions. An alternative to mean-based reference point, i.e., averaging the non-focal predictors, is the \emph{population-based} approach, discussed in detail later, which involves computing the prediction over the population of the non-focal predictors and then averaging across the values of the focal predictor \cite{hanmer2013behind}.

@@

In many applications, it usually important to report the estimates that reflect the expected values of the untransformed response. However, when dealing with nonlinear link functions, it is even harder to generate correct predictions that reflect the untransformed response due to the bias in the expected mean induced by the nonlinear transformation of the response variable. In such cases, bias correction is needed when back-transforming the predictions to the original scales. Most common approach for bias-adjustment is second-order Taylor approximation \cite{lenth2018package, duursma2003bias}. Another potential source of bias comes from additional non-focal predictors, especially in models with nonlinear link functions. Here, we describe and implement a different approach, population-based approach for bias correction.

@@

\section*{Mean-based vs population-based}

As mentioned above, the two common choices to generate predictor effects are: 1) setting the non-focal predictors to their mean -- mean-based; and 2) using the entire population of the non-focal predictors -- population-based. If the link function is nonlinear and/or the model has complex higher order interactions, the average of the predictions evaluated at the mean of the non-focal predictors and the average of predictions evaluated at the population level (and then averaged) of the non-focal predictor are not equivalent, i.e.,

@@

As stated above, in simple linear models without interactions, the effect is constant, so both approaches yield similar results. However, picking a single value, e.g., mean of the predictor, on which to draw conclusions about the effect can be problematic, unrealistic or not contained in or representative of the population. In addition, the mean-based approach fails to use every values of the non-focal predictors hence not utilizing the full potential of the information contained in the data. This may limit the inferences we can make about the entire population. In general, the mean-based approach provides the predictor effect of an average case, whereas, the population-based approach, summarizes the predictor effect over the entire population -- and in some applications, the effect of an average case might not be generalizable to the entire population, especially, if the average does not represent the population. The population-based focuses on specific observations since the prediction is first obtained for each observation and then averaged across the levels of the focal predictor.

@@

Population-based approach is not entirely foolproof. For instance, similar to the mean-based approach, in the case of continuous focal predictors, choosing the representative values of the focal predictors can be very challenging especially if the cases are not evenly distributed around the minimum and the maximum values or within some subgroups defined in the population. In addition, population-based approach can be very computationally intensive for large datasets.

@@

We now illustrate the construction of predictor effects with isolated confidence intervals and also demonstrate that the mean-based and population approaches produce different results in models with nonlinear link functions and/or additional source of potential bias. In addition, we demonstrate that population-based approach can be used to correct bias in the predictions.

@@

Mean-based and population-based approaches can produce very different results in models with nonlinear link functions with additional sources of potential bias such as additional non-focal predictor(s), random effect terms or complex interactions. To demonstrate this, we considered a two predictor binary outcome simulation described in \nameref{S2_Appendix}, such that the age effect is way smaller than that of the wealth index, and compared their predictor effect on the predicted probability of improved water quality as shown in Fig~\ref{fig:pred_bin_plots}.

\begin{figure}[!h]
\centering
\includegraphics[width=0.8\textwidth]{pred_bin_plots-figure5.pdf}
\caption{{\bf Mean-based and population-based predictor effects.} The two approaches produce relatively different predictor effects for each of the predictors but the difference is more pronounced when the effect of the particular focal predictor has no strong effect -- \emph{on the left}. In both cases, age and wealth index effect, the average predicted probability of improved water based on population-based approach is very close to the observed proportion in the simulated data. The horizontal blue, black and the red dotted lines are the average mean-based, population-based predicted probability and observed proportion of improved water quality, respectively. The vertical dotted black line represents the mean of the focal predictor, and the point at which it crosses the red dotted line represents the expected ``perfect'' prediction at the model center. The grey points are binned observations -- observed proportions of improved water quality in each bin. We would expect all the curves, horizontal, and vertical lines to cross at the model center but due the nonlinear averaging (at least in the bias corrected population-based approach) of the predictions on the response scale, this may not always be the case.} 
\label{fig:pred_bin_plots}
\end{figure}

@@

Consider the simple indirect mediation simulation described in \nameref{S3_Appendix}. We then fitted two models -- \emph{non-mediated} which models \emph{z} as a function of \emph{x} and the \emph{mediated} which models \emph{z} as a function of both \emph{x} and \emph{y}. In both cases, we compared mean-based and the population-based predictions, as shown in Fig~\ref{fig:pred_mediated_plots}.

\begin{figure}[!h]
\centering
\includegraphics[width=0.8\textwidth]{pred_mediated_plots-figure6.pdf}
\caption{{\bf Mean-based and population-based predictor effects for mediated effect.} These figures compare the mean-based and population-based predictor effect of $x$ from a mediated, $z \sim x + y$, and non-mediated, $z \sim x$, models. A: In the absence of mediator variable, both mean-based and population-based approaches predictors are identical since there is no any other additional (non-focal predictors) sources of bias. Consequently, in both approaches, the average predicted probability is very close to the observed proportion. B: When the mediator variable is included, there is no direct effect of $x$ on $z$ and as a result the predictor effect curve does not align with the observations. However, the population-based approach still closely approximates the average proportion in the simulated data. The horizontal blue, black and the red dotted lines are the average mean-based, population-based predicted probability and observed proportion, respectively. The vertical dotted black line represents the mean of the focal predictor, and the point at which it crosses the red dotted line represents the expected ``perfect'' prediction at the model center. The grey points are the binned observations.}
\label{fig:pred_mediated_plots}
\end{figure}

@@

Although commonly used \proglang{R} software packages, by default, implements mean-based approach, our simulation results demonstrated that the population-based approach has a potential of yielding results which are more consistent with the observed data. We would, therefore, argue that the use of mean-based approach should have some theoretical justification, especially in complex models.

@@

From our theoretical, methodological and simulation results, researchers using these kind of models should, in the absence of theoretical justification, report predictions based on population-based approach or at least attempt to do a comparison of the two approaches before settling on the most appropriate in answering their research question. Moreover, we provide \proglang{R} package, \pkg{vareffects}, which implements these methods and is available on github (\href{https://github.com/mac-theobio/effects}{https://github.com/mac-theobio/effects}).

@@

\begin{align}\label{sim:simple_mediate}
z_i &\sim \mathrm{Binomial(1, P(z_i = 1))} \nonumber\\
\mathrm{logit(P}(z_i = 1)) &= \eta_i \nonumber\\
\eta_i &= \beta_0 + \beta_{yz} y_i \nonumber\\
y_i &= \beta_{xy} x_i + \epsilon_y \nonumber\\
x_i &\sim \mathrm{Normal(0, 1)} \nonumber\\
\epsilon_y &\sim \mathrm{Normal(0, 1)} \nonumber\\
\beta_0 &= 5 \nonumber\\
\beta_{xy} &= 1 \nonumber\\
\beta_{yz} &= 1.5 \nonumber\\
i &= 1,\cdots, 10000
\end{align}

>>>>

In many applications which use generalized linear (mixed) models, outcome predictions or predicted probabilities are often of interest. For models that involve complex multiplicative interactions, additional non-focal predictors or nonlinear link functions, the estimated coefficient are not readily interpretable. A general way to summarize these kind of models is through predictor effects, which are not only dependent on choice of representative values of focal predictor but also sensitive to which values of non-focal predictors are chosen. The most common approach is generating predictor effects at a ``reference point'', usually the mean, of non-focal predictor, i.e., mean-based which could be considered as the effect of an ``average case'' in the population. In the presence of sources of bias such as additional non-focal predictors, nonlinear link functions, random effect terms, etc., mean-based approach generate predictor effects that are biased and may not be consistent with the observed quantities. An alternative is the whole-sample-based approach which estimates the average effect in the population. Moreover, isolated confidence intervals provide an alternative and a more clear way to describe uncertainty associated with the focal predictor of interest. In addition to theoretical and methodical comparison, using simulation, we illustrate the two approaches and show that they can produce substantially different results and that whole-sample-based approach can not only produce estimates consistent with the observed values, but also appropriate for bias correction. We also present an alternative way, isolated confidence intervals, to describe uncertainties associated with these predictions.

@@

When dealing with nonlinear link functions or models with multiple non-focal predictors, the correct predictions, for example, are even much harder to estimate. One approach is to make predictions on the transformed scale (linear predictor scale), and then back-transform to the original scale. However, the back-transformation may either result in biased predictions or requires some approximation. In particular, bias in expected mean prediction induced by nonlinear transformation of the response variable can lead inaccurate predictions. An alternative to mean-based reference point, i.e., averaging the non-focal predictors, is the \emph{whole-sample-based} approach, discussed in detail later, which involves computing the prediction over the population of the non-focal predictors and then averaging across the values of the focal predictor \cite{hanmer2013behind}.

@@

In many applications, it usually important to report the estimates that reflect the expected values of the untransformed response. However, when dealing with nonlinear link functions, it is even harder to generate correct predictions that reflect the untransformed response due to the bias in the expected mean induced by the nonlinear transformation of the response variable. In such cases, bias correction is needed when back-transforming the predictions to the original scales. Most common approach for bias-adjustment is second-order Taylor approximation \cite{lenth2018package, duursma2003bias}. Another potential source of bias comes from additional non-focal predictors, especially in models with nonlinear link functions. Here, we describe and implement a different approach, whole-sample-based approach for bias correction.

@@

\section*{Mean-based vs whole-sample-based}

As mentioned above, the two common choices to generate predictor effects are: 1) setting the non-focal predictors to their mean -- mean-based; and 2) using the entire population of the non-focal predictors -- whole-sample-based. If the link function is nonlinear and/or the model has complex higher order interactions, the average of the predictions evaluated at the mean of the non-focal predictors and the average of predictions evaluated at the population level (and then averaged) of the non-focal predictor are not equivalent, i.e.,

@@

As stated above, in simple linear models without interactions, the effect is constant, so both approaches yield similar results. However, picking a single value, e.g., mean of the predictor, on which to draw conclusions about the effect can be problematic, unrealistic or not contained in or representative of the population. In addition, the mean-based approach fails to use every values of the non-focal predictors hence not utilizing the full potential of the information contained in the data. This may limit the inferences we can make about the entire population. In general, the mean-based approach provides the predictor effect of an average case, whereas, the whole-sample-based approach, summarizes the predictor effect over the entire population -- and in some applications, the effect of an average case might not be generalizable to the entire population, especially, if the average does not represent the population. The whole-sample-based focuses on specific observations since the prediction is first obtained for each observation and then averaged across the levels of the focal predictor.

@@

Population-based approach is not entirely foolproof. For instance, similar to the mean-based approach, in the case of continuous focal predictors, choosing the representative values of the focal predictors can be very challenging especially if the cases are not evenly distributed around the minimum and the maximum values or within some subgroups defined in the population. In addition, whole-sample-based approach can be very computationally intensive for large datasets.

@@

We now illustrate the construction of predictor effects with isolated confidence intervals and also demonstrate that the mean-based and population approaches produce different results in models with nonlinear link functions and/or additional source of potential bias. In addition, we demonstrate that whole-sample-based approach can be used to correct bias in the predictions.

@@

Mean-based and whole-sample-based approaches can produce very different results in models with nonlinear link functions with additional sources of potential bias such as additional non-focal predictor(s), random effect terms or complex interactions. To demonstrate this, we considered a two predictor binary outcome simulation described in \nameref{S2_Appendix}, such that the age effect is way smaller than that of the wealth index, and compared their predictor effect on the predicted probability of improved water quality as shown in Fig~\ref{fig:pred_bin_plots}.

\begin{figure}[!h]
\centering
\includegraphics[width=0.8\textwidth]{pred_bin_plots-figure5.pdf}
\caption{{\bf Mean-based and whole-sample-based predictor effects.} The two approaches produce relatively different predictor effects for each of the predictors but the difference is more pronounced when the effect of the particular focal predictor has no strong effect -- \emph{on the left}. In both cases, age and wealth index effect, the average predicted probability of improved water based on whole-sample-based approach is very close to the observed proportion in the simulated data. The horizontal blue, black and the red dotted lines are the average mean-based, whole-sample-based predicted probability and observed proportion of improved water quality, respectively. The vertical dotted black line represents the mean of the focal predictor, and the point at which it crosses the red dotted line represents the expected ``perfect'' prediction at the model center. The grey points are binned observations -- observed proportions of improved water quality in each bin. We would expect all the curves, horizontal, and vertical lines to cross at the model center but due the nonlinear averaging (at least in the bias corrected whole-sample-based approach) of the predictions on the response scale, this may not always be the case.} 
\label{fig:pred_bin_plots}
\end{figure}

@@

Consider the simple indirect mediation simulation described in \nameref{S3_Appendix}. We then fitted two models -- \emph{non-mediated} which models \emph{z} as a function of \emph{x} and the \emph{mediated} which models \emph{z} as a function of both \emph{x} and \emph{y}. In both cases, we compared mean-based and the whole-sample-based predictions, as shown in Fig~\ref{fig:pred_mediated_plots}.

\begin{figure}[!h]
\centering
\includegraphics[width=0.8\textwidth]{pred_mediated_plots-figure6.pdf}
\caption{{\bf Mean-based and whole-sample-based predictor effects for mediated effect.} These figures compare the mean-based and whole-sample-based predictor effect of $x$ from a mediated, $z \sim x + y$, and non-mediated, $z \sim x$, models. A: In the absence of mediator variable, both mean-based and whole-sample-based approaches predictors are identical since there is no any other additional (non-focal predictors) sources of bias. Consequently, in both approaches, the average predicted probability is very close to the observed proportion. B: When the mediator variable is included, there is no direct effect of $x$ on $z$ and as a result the predictor effect curve does not align with the observations. However, the whole-sample-based approach still closely approximates the average proportion in the simulated data. The horizontal blue, black and the red dotted lines are the average mean-based, whole-sample-based predicted probability and observed proportion, respectively. The vertical dotted black line represents the mean of the focal predictor, and the point at which it crosses the red dotted line represents the expected ``perfect'' prediction at the model center. The grey points are the binned observations.}
\label{fig:pred_mediated_plots}
\end{figure}

@@

Although commonly used \proglang{R} software packages, by default, implements mean-based approach, our simulation results demonstrated that the whole-sample-based approach has a potential of yielding results which are more consistent with the observed data. We would, therefore, argue that the use of mean-based approach should have some theoretical justification, especially in complex models.

@@

From our theoretical, methodological and simulation results, researchers using these kind of models should, in the absence of theoretical justification, report predictions based on whole-sample-based approach or at least attempt to do a comparison of the two approaches before settling on the most appropriate in answering their research question. Moreover, we provide \proglang{R} package, \pkg{vareffects}, which implements these methods and is available on github (\href{https://github.com/mac-theobio/effects}{https://github.com/mac-theobio/effects}).

@@

\begin{align}\label{sim:simple_mediate}
z_i &\sim \mathrm{Binomial(1, P(z_i = 1))} \nonumber\\
\mathrm{logit(P}(z_i = 1)) &= \eta_i \nonumber\\
\eta_i &= \beta_0 + \beta_{xz} x_i + \beta_{yz} y_i \nonumber\\
y_i &= \rho x_i + \sqrt{1-\rho^2} y_y \nonumber\\
x_i &\sim \mathrm{Normal(0, 1)} \nonumber\\
y_y &\sim \mathrm{Normal(0, 1)} \nonumber\\
\rho &= 0.8 \nonumber\\
\beta_0 &= 5 \nonumber\\
\beta_{xz} &= 0.2 \nonumber\\
\beta_{yz} &= 1.5 \nonumber\\
i &= 1,\cdots, 10000
\end{align}
